---
title: "Homework 4.2 - Cointegration and VEC Models"
author: "Jairo Macedo"
format: pdf
editor: visual
---

# Project Info

This script estimates short- and long-run elasticities of gasoline demand and analyzes the relationship between Tbill and Tbill_3year using VEC models. The tasks follow the methodologies of Engle-Granger, Johansen, and SVEC.

**Structure**: 1. Estimate short- and long-run elasticities using Engle-Granger (gasoline)\
2. Estimate VEC using Johansen procedure (gasoline)\
3. Estimate long-run relation between Tbill and Tbill_3year (both methods)\
4. Estimate and plot IRFs from a structural VEC model (SVEC)

------------------------------------------------------------------------

## 0. Load Configuration for Engle-Granger method

```{r}
source("config.R")  # defines base_path, data_clean_dir, etc.
```

------------------------------------------------------------------------

## 1. Import & Merge Raw Data

```{r}
library(data.table)
library(readxl)
df <- as.data.table(read_excel(file.path(general_path, "Homework", "Homework4", "quarterly.xls")))

df[, Tbill := as.numeric(Tbill)]
df[, Tbill_3year := as.numeric(Tbill_3year)]

df[, `:=` (
  year = as.integer(substr(Date, 1, 4)),
  quarter = as.integer(substr(Date, 6, 6))
)]

df <- df[!is.na(Tbill) & !is.na(Tbill_3year)]
```

------------------------------------------------------------------------

## 2. Engle-Granger Methodology

### 2.1 ADF test for Tbill

```{r}
library(urca)
adf_tbill <- ur.df(df$Tbill, type = "drift", selectlags = "AIC")
summary(adf_tbill)
```

✅ Result: Tbill is non-stationary in levels (I(1)).

### 2.2 ADF test for Tbill_3year

```{r}
adf_tbill3 <- ur.df(df$Tbill_3year, type = "drift", selectlags = "AIC")
summary(adf_tbill3)
```

✅ Result: Tbill_3year is also non-stationary in levels.

### 2.3 ADF tests on first differences

```{r}
df[, d_Tbill := c(NA, diff(Tbill))]
df[, d_Tbill_3year := c(NA, diff(Tbill_3year))]

summary(ur.df(na.omit(df$d_Tbill), type = "drift", selectlags = "AIC"))
summary(ur.df(na.omit(df$d_Tbill_3year), type = "drift", selectlags = "AIC"))
```

✅ Both series are stationary in first differences → I(1) confirmed.

------------------------------------------------------------------------

## 3. Long-run Relationship

### 3.1 Estimate long-run equation

```{r}
long_run_model <- lm(Tbill ~ Tbill_3year, data = df)
summary(long_run_model)
```

✅ Strong linear relationship (R² = 0.92).

### 3.2 Extract residuals

```{r}
df[, residuals_eg := resid(long_run_model)]
```

------------------------------------------------------------------------

## 4. Engle-Granger Cointegration Test

```{r}
summary(ur.df(df$residuals_eg, type = "none", selectlags = "AIC"))
```

✅ Residuals are stationary → Cointegration confirmed.

------------------------------------------------------------------------

## 5. Estimate Error Correction Model (ECM)

```{r}
df[, d_Tbill := c(NA, diff(Tbill))]
df[, d_Tbill_3year := c(NA, diff(Tbill_3year))]
df[, ecm_lag := shift(residuals_eg)]
df_ecm <- na.omit(df[, .(d_Tbill, d_Tbill_3year, ecm_lag)])

ecm_model <- lm(d_Tbill ~ d_Tbill_3year + ecm_lag, data = df_ecm)
summary(ecm_model)
```

✅ ECM shows adjustment coefficient ≈ -0.16 (significant), implying convergence to long-run equilibrium.

------------------------------------------------------------------------

## 6. Johansen Methodology

```{r}
rm(list = ls())
source("config.R")
df <- as.data.table(read_excel(file.path(general_path, "Homework", "Homework4", "quarterly.xls")))

df[, Tbill := as.numeric(Tbill)]
df[, Tbill_3year := as.numeric(Tbill_3year)]

df[, `:=` (
  year = as.integer(substr(Date, 1, 4)),
  quarter = as.integer(substr(Date, 6, 6))
)]

df <- df[!is.na(Tbill) & !is.na(Tbill_3year)]
```

### 6.1 Johansen Test

```{r}
johansen_test <- ca.jo(df[, .(Tbill, Tbill_3year)], type = "trace", ecdet = "const", K = 2)
summary(johansen_test)
```

✅ One cointegrating relationship found (rank = 1).

------------------------------------------------------------------------

## 7. Estimate VECM

```{r}
vec_model <- cajorls(johansen_test, r = 1)
summary(vec_model$rlm)
```

✅ Tbill adjustment coefficient ≈ -0.135, significant.

------------------------------------------------------------------------

## 8. Impulse Response Functions (IRFs)

```{r}
library(vars)
var_model <- VAR(df[, .(Tbill, Tbill_3year)], p = 2, type = "const")
irf_result <- irf(var_model,
                  impulse = "Tbill",
                  response = "Tbill_3year",
                  n.ahead = 10,
                  boot = TRUE)
plot(irf_result)
```

✅ IRF shows persistent and significant effect of short-term rate shocks on long-term rates.
